Torrential Twitter? Measuring the Severity of Harassment when Canadian Female Politicians Tweet about Climate Change

### 2024-02-12 Outputs/Paper/Publication_version/paper_for_pub.qmd
update

To gain a deeper understanding of the intersection of climate change denialism and online gendered-harassment faced by Canadian female politicians, I collected and manually coded replies to Tweets from DPM and MP for University--Rosedale, Hon. Chrystia Freeland (from the Liberal Party), MP for Saanich--Gulf Islands, Elizabeth May (Co-Leader of the Green Party), and MP for Victoria, Laurel Collins (from the New Democratic Party) that specifically discuss climate change and environmental policy. Tweets were coded using a seven-point scale to measure the severity of harassment and a six-point scale to categorize the types of accounts behind the replies. I then built a Bayesian hierarchical model to explore the relationship between the type of account, whether a user was verified through Twitter's subscription program, and which female politician the reply was directed toward to see whether certain combinations of these variables are more likely to lead to a harassing reply.

@@

The notion of having a public sphere for political deliberation is as old as Athenian democracy itself [@zizi2002]. In its simplest form, the public sphere is a place where "...society engages in critical public debate" [@habermas1989, p. 52] and meaning is "...generated, circulated, contested, and reconstructed" [@fraser1995, p. 287]. Although Jürgen Habermas wrote about the public sphere in the mid-twentieth century, eighteenth century bourgeois society served as the basis for his theory, overlooking how elites tried to influence and dominate the form political discourse took among other socio-economic classes [@mediatingmisogyny; @megarry2014]. Through Habermas' conceptualization, the public sphere intends to magnify the voices of the political elites (traditionally white men), while diminishing and even vilifying the voices of minorities -- traits which have been transposed into the digital public sphere [@mediatingmisogyny].

The past few decades have led to the structural transformation and fragmentation of the public sphere, moving from having a singular, "official" public sphere to a multitude of parallel smaller ones, called "subaltern counterpublics" and networked publics mediated by technological affordances [@brunshighfield; @fraser1991; @boyd2010]. Technology reorganized the flow of information in society, distinguishing networked publics from subaltern counterpublics and the "official" public. Discourse traditionally happened in mediated spaces, but the rise of social media allows it to happen everywhere, theoretically breaking down barriers of who has access to the walled garden of democratic discourse [@mediatingmisogyny; @trifiro]. Despite the potential for the formation of new publics online, including issue publics driven by shared policy interests such as climate change, the leading voices in networked publics largely continue to be white, male, Christian, and have free time (often dictated by patriarchal norms) to actively contribute to discussions [@bruns2023; @mantilla2013; @degeus; @tweetingclimate]. Networked publics use trolling and harassment to undermine the authority of female politicians whose actions do not fit traditional gender roles and conceptions of power [@degeus; @mantilla2015; @megarry2014]. Consequently, women have and continue to be on the margins of the public sphere, both in physical spaces like the House of Commons and online, as underscored by an @ipu global survey which found that 82% of female politicians reported being subjected to psychological violence (including harassment, threats, and sexist comments). In Canada, specific examples of on-going marginalization can be seen through the lived experiences of women in politics, with only one female Prime Minister to date (the Rt. Hon. Kim Campbell) and 30% of federal MPs being female [@degeus; @hoc].

The original mindset for the development of the internet in the second half of the twentieth century was one of utopianism and democratic renewal -- a new space for civil conversations [@zizi2002; @zizi2004]. However, early 2000s political communication theorists often overlooked the aims and agendas of the people who engineered these technologies [@zizi2004]. @megarrybook adds that the internet was seen as the "last frontier" for male domination, much like the classic image of the American Wild West. The behaviours and norms of the Wild West continue to play out through the actions of male tech CEOs, such as Elon Musk, and through the way platforms are governed. While often lacking in application, social media platforms including Twitter have slowly adopted rigorous content moderation and community safety policies [@overtime].

@@

Diverging scholarly viewpoints exist when operationalizing and applying the theory of deliberative democracy. Some scholars view the theory in rigid terms, evaluating all characteristics and scenarios, such as political parties, systems, and institutions on a "yes" or "no" basis [@sokolon]. However, as @rossini suggests, the theory of deliberative democracy can be viewed on a scale -- looking at the extent in which healthy democratic speech exists in specific contexts. In our increasingly digitized democracy, we cannot overlook or discount how algorithms and platform affordances are building the networks and contexts in which influential democratic deliberation takes place [@bruns2023].

@@

All Twitter data was collected throughout the month of October 2023 using the statistical programming software R [@r] and functions from `voson.tcn` [@gertzel2021] and the `tidyverse` [@rTidyverse]. Paid access to Twitter's Basic v2 Application Programming Interface (API) endpoint was obtained before using the advanced search feature to determine which Tweets I wanted to download via the API. My sample includes any original Tweets written in English (no quote or Retweets) that originated on the accounts of DPM Freeland (\@cafreeland), MP May (\@elizabethmay), and MP Collins (\@laurel_bc) that were posted between September 1, 2022, and October 31, 2023. These three politicians were selected based on their pro-climate change stances, evidence of harassing Tweets, and diversity in regional representation and political affiliation. Only 6% of female MPs identify as a visible minority or Indigenous and do not frequently discuss climate change on Twitter, which is why my sample contains all white MPs [@hoc]. Tweets also had to the contain the word "climate," "environment," and/or "environmental" (in text or as a hashtag). For threaded Tweets sent by the politicians, the first Tweet of the thread had to contain at least one of the key words and the overall focus of the thread had to be on climate change and/or environmental policy. Key words were also checked for applicability, given the use of words such as "climate" and "environment" to reference geopolitics and economic situations and not actual climate or environmental policy. Within the parameters of the sample, DPM Freeland's Tweets were from October 2022 to October 2023 and had either the key word "climate" or "environment". MP May's Tweets were from September 2022 to October 2023 and included the words "climate," "environment," and "environmental". Lastly, MP Collins' Tweets were from October 2022 to October 2023 and included the words "climate," "environment," and "environmental."

@@

To ensure validity, an undergraduate research assistant coded a random selection of 10% (413) of the total number of Tweets in the sample. Intercoder reliability was calculated using the R package `irr` [@irr] to check the extended percentage agreement among coders. Acceptable levels of agreement were reached, with agreement for severity of harassment at 88% and type of account at 76% [@lacy2015].

@@

This model investigates the combination of the type of account a user has, whether they are verified, and which politician they are replying to in order to predict if they are more likely to send a harassing Tweet. Historically, accounts with the blue checkmark/tick were regarded as credible individuals and/or organizations, sharing legitimate information [@unverified]. However, since Musk introduced changes to Twitter's verification policy in April 2023, the @CCDHreport found that 99% of online hate was spread by malicious users who paid to be verified, allowing for circumvention of content moderation rules. Bayesian hierarchical modeling using the statistical programming language R [@r], employing `rstanarm` [@rstanarm], `marginaleffects` [@marginal], and `modelsummary` [@modelsummary] can help predict and further investigate this relationship. The model was fit in a Bayesian framework using `rstanarm` [@rstanarm]. For the priors, I followed the standard weakly informed prior distributions by using the normal definition with mean 0 and standard deviation 2.5 as used in the `rstanarm` package.

@@

harassment_model_predictions |>
  ggplot(aes(x = harassment_binary, y = estimate, color = name_of_politician)) +
  geom_jitter(width = 0.2, height = 0.0, alpha = 0.3) +
  labs(
    x = "Harassment",
    y = "Estimated probability that a Tweet contains harassment",
    color = "Name of Politician"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust = 1, size = 8)) + 
  theme(axis.text.y = element_text(size = 8)) + 
  theme(axis.title.y = element_text(size = 8)) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9))

@@

# Acknowledgments 
The author(s) thank(s) Rohan Alexander for the helpful suggestions and feedback on drafts of this paper and Elizabeth Dubois for guidance during the early conception of this study. The author(s) also thank(s) Sarah Xie and Allyson Cui for their research assistance.

# Funding 
The author(s) disclose(s) receipt of the following financial support for the research, authorship, and/or publication of this article: This study was funded by the Social Sciences and Humanities Research Council of Canada (grant ID: 766-2022-0719) and the Government of Ontario, *2023-2024 Ontario Graduate Scholarship*.

# Declaration of Conflicting Interests
The author(s) declare(s) that there is no conflict of interest.

>>>>

To gain a deeper understanding of the intersection of climate change denialism and online gendered-harassment faced by Canadian female politicians, I collected and manually coded replies to Tweets from DPM and MP for University--Rosedale, Hon. Chrystia Freeland (from the Liberal Party), MP for Saanich--Gulf Islands, Elizabeth May (Co-Leader of the Green Party), and MP for Victoria, Laurel Collins (from the New Democratic Party) that specifically discuss climate change and environmental policy. Tweets were coded using a seven-point scale to measure the severity of harassment and a six-point scale to categorize the types of accounts behind the replies. I then built a Bayesian hierarchical model to explore the relationship between the type of account, whether a user was verified through Twitter's subscription program, and which female politician the reply was directed toward, to see whether certain combinations of these variables are more likely to lead to a harassing reply.

@@

The notion of having a public sphere for political deliberation is as old as Athenian democracy itself [@zizi2002]. In its simplest form, the public sphere is a place where "...society engages in critical public debate" [@habermas1989, p. 52] and meaning is "...generated, circulated, contested, and reconstructed" [@fraser1995, p. 287]. Although Jürgen Habermas wrote about the public sphere in the mid-twentieth century, eighteenth century bourgeois society served as the basis for his theory, overlooking how elites tried to influence and dominate the form political discourse took among other socio-economic classes [@mediatingmisogyny; @megarry2014]. Through Habermas' conceptualization, the public sphere tends to magnify the voices of the political elites (traditionally white men), while diminishing and even vilifying the voices of minorities -- traits which have been transposed into the digital public sphere [@mediatingmisogyny].

The past few decades have led to the structural transformation and fragmentation of the public sphere, moving from having a singular, "official" public sphere to a multitude of parallel smaller ones, called "subaltern counterpublics" and networked publics mediated by technological affordances [@brunshighfield; @fraser1991; @boyd2010]. Technology reorganized the flow of information in society, distinguishing networked publics from subaltern counterpublics and the "official" public. Discourse traditionally happened in mediated spaces, but the rise of social media allows it to happen everywhere, theoretically breaking down barriers of who has access to the walled garden of democratic discourse [@mediatingmisogyny; @trifiro]. Despite the potential for the formation of new publics online, including issue-specific publics driven by shared policy interests such as climate change, the leading voices in networked publics largely continue to be white, male, Christian, and have free time (often dictated by patriarchal norms) to actively contribute to discussions [@bruns2023; @mantilla2013; @degeus; @tweetingclimate]. Networked publics use trolling and harassment to undermine the authority of female politicians whose actions do not fit traditional gender roles and conceptions of power [@degeus; @mantilla2015; @megarry2014]. Consequently, women have and continue to be on the margins of the public sphere, both in physical spaces like the House of Commons and online. This is underscored by an @ipu global survey which found that 82% of female politicians reported being subjected to psychological violence (including harassment, threats, and sexist comments). In Canada, specific examples of on-going marginalization can be seen through the lived experiences of women in politics, with only one female Prime Minister to date (the Rt. Hon. Kim Campbell) and only 30% of federal MPs being female [@degeus; @hoc].

The original mindset for the development of the internet in the second half of the twentieth century was one of utopianism and democratic renewal -- a new space for civil conversations [@zizi2002; @zizi2004]. However, early 2000s political communication theorists often overlooked the aims and agendas of the people who engineered these technologies [@zizi2004]. @megarrybook adds that the internet was seen as the "last frontier" for male domination, much like the classic image of the American Wild West. The behaviours and norms of the Wild West continue to play out through the actions of male tech CEOs, such as Elon Musk, and through the way platforms are governed. While often lacking in application, social media platforms including Twitter have tended to slowly adopt rigorous content moderation and community safety policies [@overtime].

@@

Diverging scholarly viewpoints exist when operationalizing and applying the theory of deliberative democracy. Some scholars view the theory in rigid terms, evaluating all characteristics and scenarios, such as political parties, systems, and institutions on a "yes" or "no" basis [@sokolon]. However, as @rossini suggests, the theory of deliberative democracy can be viewed on a scale -- looking at the extent to which healthy democratic speech exists in specific contexts. In our increasingly digitized democracy, we cannot overlook or discount how algorithms and platform affordances are building the networks and contexts in which influential democratic deliberation takes place [@bruns2023].

@@

All Twitter data was collected using the statistical programming software R [@r] and functions from `voson.tcn` [@gertzel2021] and the `tidyverse` [@rTidyverse]. Paid access to Twitter's Basic v2 Application Programming Interface (API) endpoint was obtained before using the advanced search feature to determine which Tweets I wanted to download via the API. My sample includes any original Tweets written in English (no quote or Retweets) that originated on the accounts of DPM Freeland (\@cafreeland), MP May (\@elizabethmay), and MP Collins (\@laurel_bc) that were posted between September 1, 2022, and October 31, 2023. These three politicians were selected based on their pro-climate change stances, evidence of harassing Tweets, and diversity in regional representation and political affiliation. Only 6% of female MPs identify as a visible minority or Indigenous and do not frequently discuss climate change on Twitter, which is why my sample contains only white MPs [@hoc]. Tweets also had to the contain the word "climate," "environment," and/or "environmental" (in text or as a hashtag). For threaded Tweets sent by the politicians, the first Tweet of the thread had to contain at least one of the key words and the overall focus of the thread had to be on climate change and/or environmental policy. Key words were also checked for applicability, given the use of words such as "climate" and "environment" to reference geopolitics and economic situations and not actual climate or environmental policy. Within the parameters of the sample, all Tweets were from September 2022 to October 2023. DPM Freeland's Tweets had either the key word "climate" or "environment." MP May and MP Collins' Tweets included the words "climate," "environment," and "environmental."

@@

To ensure validity, an undergraduate research assistant coded a random selection of 10% (413) of the total number of Tweets in the sample. Intercoder reliability was calculated using the R package `irr` [@irr] to check the extended percentage agreement among coders. Acceptable levels of agreement were reached according to @lacy2015, with agreement for severity of harassment at 88% and type of account at 76%.

@@

This model investigates the combination of the type of account a user has, whether they are verified, and which politician they are replying to in order to predict if they are more likely to send a harassing Tweet. Historically, accounts with the blue checkmark/tick were regarded as credible individuals and/or organizations, sharing legitimate information [@unverified]. However, since Musk introduced changes to Twitter's verification policy in April 2023, the @CCDHreport found that 99% of online hate was spread by malicious users who paid to be verified, allowing for circumvention of content moderation rules. Bayesian hierarchical modeling using the statistical programming language R [@r], employing `rstanarm` [@rstanarm], `marginaleffects` [@marginal], and `modelsummary` [@modelsummary] can help predict and further investigate this relationship. The model was fit using `rstanarm` [@rstanarm]. For the priors, I followed the standard weakly informed prior distributions by using the normal definition with mean 0 and standard deviation 2.5 as used in the `rstanarm` package.

@@

harassment_model_predictions |>
  ggplot(aes(x = harassment_binary, y = estimate, color = name_of_politician)) +
  geom_jitter(width = 0.2, height = 0.0, alpha = 0.3) +
  labs(
    x = "Contains Harassment",
    y = "Estimated probability that a Tweet \ncontains harassment",
    color = "Name of Politician"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust = 1, size = 10)) + 
  theme(axis.text.y = element_text(size = 10)) + 
  theme(axis.text.x = element_text(size = 10)) +
  theme(axis.title.y = element_text(size = 12)) +
  theme(axis.title.x = element_text(size = 12)) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9))

@@

# Acknowledgments 
The author thanks Rohan Alexander for the helpful suggestions and feedback on drafts of this paper and Elizabeth Dubois for guidance during the early conception of this study. The author also thanks Sarah Xie and Allyson Cui for their research assistance.

# Funding 
The authors discloses receipt of the following financial support for the research, authorship, and/or publication of this article: This study was funded by the Social Sciences and Humanities Research Council of Canada (grant ID: 766-2022-0719) and the Government of Ontario, *2023-2024 Ontario Graduate Scholarship*.

# Declaration of Conflicting Interests
The author declares that there is no conflict of interest.
